# The Starships {#ships}

## The Data

```{r load_starships_data, warning = FALSE}
# load and format data
starships <- data.frame(t(list.cbind(sw_starships))) %>%
  # convert the star ways starship dataset to a tibble
  as_tibble() %>%
  # clean column names
  janitor::clean_names() %>%
  # select only those columns between name and starship (inclusive)
  select(name:starship_class) %>%
  # mutate columns that are lists to non-lists
  mutate_if(is.list, ~(unlist(.))) %>%
  # mutate all columns, replacing "unknown" with "NA"  
  mutate_all(~(na_if(., "unknown"))) %>%
  # convert numeric columns to numerics
  mutate_at(vars(cost_in_credits,
                 length,
                 crew,
                 passengers,
                 cargo_capacity,
                 hyperdrive_rating,
                 mglt), ~(as.numeric(.))) %>%
  # extract the max_atmosphering_speed and convert starship_class to a lowercase
  # factor
  mutate(max_atmosphering_speed = parse_number(max_atmosphering_speed),
         starship_class         = as.factor(tolower(starship_class))) %>%
  # seperate consumables into a value and unit (e.g., days, weeks, months, 
  # years)
  separate(consumables, c("con_n", "con_unit"), sep = " ") %>%
  # convert units to numeric values and multiply with the consumable value to 
  # provide the amount of consumable in the total number of days
  mutate(con_unit    = case_when(con_unit == "day"   ~ "1",
                                 con_unit == "week"  ~ "7",
                                 con_unit == "month" ~ "30",
                                 con_unit == "years" ~ "365"),
         consumables = as.numeric(con_n) * as.numeric(con_unit)) %>%
  # retain only the new consumables column (i.e., drop con_n and con_unit)
  select(-starts_with("con_"))

# look at first ten rows of tidied dataet
head(starships, n = 10L) 
```

## The Visualization

```{r visualization_2}
# generate random x, y, and transparency data to use for plotting stars
stars_data <- tibble(x     = sample(seq(from =  0, to =  4, by = .0001), 500),
                     y     = sample(seq(from =  0, to = 80, by = .0001), 500),
                     trans = sample(seq(from = .1, to = .9, by = .0001), 500))

# starships plot
starships %>%
  # group by starship_class
  group_by(starship_class) %>%
  # drop rows that have an NA for length 
  filter(!is.na(length)) %>%
  # calculate mean length and standard deviation, as well as number of ships in 
  # that class
  summarize(len_m  = mean(length),
            len_sd = sd(length),
            len_n  = n()) %>%
  # retain classes that have more than one ship
  filter(len_n > 1) %>%
  # convert starship_class to title case and wrap multi-word levels
  mutate(starship_class = str_to_title(starship_class),
         starship_class = gsub(" ", " \n", starship_class)) %>%
  # calculate standard error
  mutate(len_se = len_sd / sqrt(len_n)) %>%
  # plot starship class v starship length
  ggplot(aes(x = fct_reorder(starship_class, desc(len_m)), y = len_m)) +
    # add error bars
    geom_errorbar(aes(ymin = len_m - (len_se * qnorm(.975)), 
                      ymax = len_m + (len_se * qnorm(.975))), 
                  width = .5, colour = "grey70", lwd = 1.25) +
    geom_errorbar(aes(ymin = len_m - (len_se * qnorm(.975)), 
                      ymax = len_m + (len_se * qnorm(.975))), 
                  width = .5, colour = "grey30", lwd = 1) +
    # add stars 
    geom_point(data = stars_data, aes(x = x, y = y, alpha = trans), 
               colour = "white",
               size = .001) +
    # plot ship body points
    geom_point(size = 6.70, colour = "grey70") +
    geom_point(size = 6.40, colour = "grey30") +
    geom_point(size = 4.40, colour = "grey10") +
    geom_point(size = 3.40, colour = "grey60", shape = 8) +
    geom_point(size = 4.40, colour = "grey60", shape = 1) +
    geom_point(size = 2.00, colour = "grey60") +
    geom_point(size = 1.50, colour = "grey10") +
    geom_point(size = 0.20, colour = "grey60", 
               position = position_nudge(x = -0.09, y = -0.5)) +
    geom_point(size = 0.20, colour = "grey60", 
             position = position_nudge(x = -0.09, y = 0.5)) +

    geom_point(size = 0.10, colour = "grey10", 
               position = position_nudge(x = -0.09, y = -0.5)) +
    geom_point(size = 0.10, colour = "grey10", 
             position = position_nudge(x = -0.09, y = 0.5)) +
    # flip the coordinates
    coord_flip() +
    # add labels
    labs(title    = "Average Length of Starship by Ship Class",
         subtitle = "Mean length indicated by ship body position\nConfidence interval (95%) indicated by wing span",
         y        = "Average Length in Meters",
         x        = "Class",
         caption  = "Data from https://swapi.co.") + 
    # set y-axis limits and do not show value beyond limits of y-axis
    scale_y_continuous(limits = c(0,80), 
                       expand = expand_scale(mult = c(0, 0))) + 
    # customize using my theme
    my_theme() +
    theme(plot.margin  = unit(c(1, 1, 1, 1), "cm"),
          axis.title.y = element_text(vjust = 5))
```

#### Idea

Points and standard error bars have always looked like Tie Fighters to me, so I decided I would try to plot the length of a starship against the starship class. As would be expected yachts are the largest ships with assult starfighters being slightly larger than basic starfighters. Although I am quite happy with the plot, I believe the fonts and text elements could be greatly improved. Representing all ships as Tie Fighters also may be misleading. 

#### Intended Audience

Again, the intended audience for this visualization are people who like Star Wars. To appreciate the error bars some knowledge of statistics would likely be needed, as well.  

#### Intended Message

That starship lengths differ by class and that the length of a yacht starship is far more variable than other starship classes. 