# The Planets {#planets}

```{r planets_initial_setup, include = FALSE}
# load necessary packages (if you do not have them, you will need to install them first)
library(tidyverse) # for dataframe manipulation and plotting
library(magrittr) # for updating an object that is piped into a function (using `%<>%`)
library(here) # for finding the project folder
library(repurrrsive) # contains the star wars datasets
library(rlist) # for binding all lit elements by column (using `list.cbind`)
library(ggrepel) # for adding geom_text labels that do not overlap
library(gganimate) # for animating the plot

# add gravitational constant (i.e., 6.67408 * 10^-11)
grav_constant <- 6.67408 * (10 ^ -11)

# load bespoke functions from the functions folder
source(here("functions/my_theme.R")) # load my_theme
source(here("functions/geom_saberhilt.R")) # load the function for creating the hilts of lightsabers
source(here("functions/geom_saberblade.R")) # load the function for creating the blades of lightsabers

# set figure output dimensions
knitr::opts_chunk$set(fig.width  = 12, 
                      fig.height = 8)

# how to build this book
# 1. install the `bookdown` packages: install.packages("bookdown")
# 2. Navigate to the RStudio pane that has the environment in it. There should be the following tabs: (1) "Environment", (2) "History", (3) "Connection", (4) "Build", and (5) "Git". Select "Build" and click the button that says "Build Book"
```


### The Data

```{r load_planets_data, warning = FALSE}
# load and format data
planets <- data.frame(t(list.cbind(sw_planets))) %>%
  # convert star wars planets dataset to a tibble
  as.tibble() %>%
  # retain only columns between name and population (inclusive)
  select(name:population) %>%
  # mutate columns that are lists to non-lists
  mutate_if(is.list, ~(unlist(.))) %>%
  # mutate all columns, replacing "unknown" with "NA"  
  mutate_all(~(na_if(., "unknown"))) %>%
  # convert numeric columns to type numeric
  mutate_at(vars(rotation_period,
                 orbital_period,
                 diameter,
                 surface_water,
                 population),
            ~(as.numeric(.))) %>%
  # calculate values of interest
  mutate(
         # make surface water a proportion
         surface_water   = surface_water / 100,
         # calculate population in millions
         population_mils = population / 1000000,
         # calculate a planet's distance from the sun (Kepler's 3rd law)
         au              = (((orbital_period / 365) ^ 2) ^ (1/3)),
         # convert gravity to a number
         gravity         = parse_number(as.character(gravity)),
         # calculate the force of gravity
         g               = gravity * 9.8, 
         # calculate radius
         radius          = diameter / 2, 
         # calculate planet mass (ish)
         mass            = (g * (radius^2)) / grav_constant,
         # calculate planet area
         area            = pi * (radius^2),
         # create column indicating whether the planet is in the solar system
         solar_system    = rep(FALSE, n()),
         # create a new variable indicating whether a planet was in the original 
         # trilogy
         orig_tril = case_when(name == "Alderaan" |
                               name == "Tatooine" |
                               name == "Yavin"    |
                               name == "Yavin 4"  ~ "new_hope",
                               name == "Bespin"   |
                               name == "Dagobah"  |                     
                               name == "Hoth"     ~ "empire_strikes",    
                               name == "Endor"    ~ "return_jedi",
                               TRUE ~ NA_character_))

# create dataset for planets in the solar system
sol_sys <- tibble(
  name           = c("Mercury", "Venus", "Earth", "Mars", "Jupiter"),
  orbital_period = c(       88,     225,     365,    687,      4330),
  au             = c(      .39,     .72,       1,   1.52,      5.20),
  solar_system   = c(     TRUE,    TRUE,    TRUE,   TRUE,      TRUE),
  radius         = c(   2439.5,    6052,    6378,   3396,     71492)
)

#combine the star war's planets and solar system data into one dataset
planets <- bind_rows(planets, sol_sys)

# look at first ten rows of tidied dataet
head(planets, n = 10L) 
```


### The Visualization
```{r visualization_3, fig.height = 12, warning = FALSE}
# create function for calculating x axis locations during the time it takes
# bespin to make one rotation. There has to be enough slices to obscure the
# geom_points jumping back at the end of Bespin's year
orbit_slice <- function(planet = NULL) {
  # grab orbital period of planet of interest
  temp <- planets$orbital_period[which(planets$name == planet)]
  # calculate 511 x-axis point for that planet throughout Bespin' year
  rep(c(seq(from       = 0, 
            to         = 5110, 
            length.out = (5100 * (temp / 5110))), 
        NA),
      length.out = 5100)
}

# calculate locations on the x-axis for each planets during the time it takes
# bespin to make one rotation. 
orbital_degrees <- tibble(
  "Alderaan"   = orbit_slice("Alderaan"),
  "Hoth"       = orbit_slice("Hoth"),
  "Dagobah"    = orbit_slice("Dagobah"),
  "Bespin"     = orbit_slice("Bespin"),
  "Endor"      = orbit_slice("Endor"),
  "Tatooine"   = orbit_slice("Tatooine"),
  "Mercury"    = orbit_slice("Mercury"),
  "Venus"      = orbit_slice("Venus"),
  "Earth"      = orbit_slice("Earth"),
  "Mars"       = orbit_slice("Mars"),
  "Jupiter"    = orbit_slice("Jupiter"),
  "days_count" = 1:5100
)

# gather days into a column and the planet names into another column
orbital_degrees %<>%
  gather(key = "name", value = "days", -days_count)

# planets plot
planets_plot <- planets %>%
  # select only variables of interest
  select(name, au, radius, orbital_period, orig_tril, solar_system) %>%
  # drop planets that do not have (1) a name, (2) a distance from the sun (i.e.
  # an AU), (3) an AU less than 2, and (4) a radius greater than 0.
  filter(!is.na(name),
         !is.na(au),
         !is.na(orbital_period),
         !is.na(orig_tril) | solar_system == TRUE,
         # au < 2,
         radius != 0.00) %>%
  left_join(orbital_degrees, by = "name") %>%
  #drop mercury and alderaan
  filter(name != "Mercury", # mercury for orbiting too quickly
         name != "Alderaan") %>% # alderaan for esssentially being earth 2.0
  # plot the planets distance from the sun against their arbitrary degree
  ggplot(aes(x = days, y = au, color = solar_system)) +
    # add labels indicating y-axis units and add star and earth
    geom_text(aes(x =   0,          y = 0, label = "STAR"), 
              color = "deeppink3", 
              size  = 5) +
    geom_text(aes(x = 5110 * 3 / 4, y = 2, label = "2 AU"), 
              color = "deeppink3", 
              size  = 5) +
    geom_text(aes(x = 5110 * 3 / 4, y = 4, label = "4 AU"), 
              color = "deeppink3", 
              size  = 5) +
    geom_text(aes(x = 5110 * 3 / 4, y = 6, label = "6 AU"), 
              color = "deeppink3", 
              size  = 5) +
    # add label for describing 1 AU
    annotate(geom  =                   "label", 
             x     =             5110 * 1 / 2,
             y     =                        7,
             label = "1 AU = ~150,000,000 km", 
             color =              "deeppink2",
             fill  =                  "black",
             size  =                        5) +
    # add planet points; map size of the point to planet radius
    geom_point(aes(size = radius), alpha = .6) +
    # add planet text; also suppress its legend
    geom_text(aes(label   = name), 
              position    = position_nudge(y = 1.1), 
              alpha       = .7,
              size        =  5,
              show.legend = FALSE) +
    # make the coordinate polar
    coord_polar(theta = "x") +
    # add transition according to days
    transition_time(days_count) +
    # set the limit of size values
    scale_size_continuous(range = c(.5, 10)) +
    # drop x-axis grid marks and breaks
    scale_x_continuous(breaks = NULL) +
    # set 2-axis limit
    scale_y_continuous(limits = c(0, 7)) +
    # set the colours
    scale_colour_manual(name   = "Star System",
                        values = c("#02fff6", "#ff7b00"),
                        labels = c("Solar System", "Star Wars")) +
    # add labels
    labs(title    = "Distance from Host Star and Orbital Period",
         subtitle =                      "Days: {frame_time}", 
         caption  =                "Data from https://swapi.co.") +
    # customize using my theme
    my_theme() +
    # drop axis text and axis titles and center title, subtitle, and caption
    theme(plot.title      = element_text(hjust = 0.5),
          plot.subtitle   = element_text(hjust = 0.5),
          plot.caption    = element_text(hjust = 0.5),
          axis.text       = element_blank(),
          axis.title      = element_blank(),
          plot.margin     = unit(c(1, 5, 1, 7), "cm"),
          legend.position = "right") +
    # drop radius legend and change the size of the legend points
    guides(size = FALSE,
           colour = guide_legend(override.aes = list(size  = 5,
                                                     alpha = 1)))

# animate plot with a one-minute duration and 24 frames per second
animate(planets_plot, 
        duration = 3,
        fps      = 1)

#5548
``` 

#### Idea

As interesting as I found the data in the first two plots, I think several plots could be devoted to the planets dataset. Specifically, the visualizations could be framed around the common theme of astronomy, serving as an instructive tool for teaching concepts from astronomy (e.g., Keppler's third law, calculations of gravitational force). The present plot shows the relationship between a planet's surface water and its distance from the sun. I would like to add a solar system dataset to the Starwar's planets dataset and use facetwrap to provide a side by side comparison of fictional and non-fictional planets. 

#### Intended Audience

Fans of Star Wars or anyone who has an interest in astronomy or phyics.

#### Intended Message

Planets farther from the sun seem to have more surface water, which seems to be consistent with our solar system. 
