---
title: "edld610_catd_finalproject_proposal"
author: "Cameron Kay"
date: "2/3/2019"
output: html_document
---

```{r load_packages, include = FALSE}
# load necessary packages
library(tidyverse)
library(magrittr)
library(repurrrsive)
library(rlist)

# add gravitational constant
grav_constant <- 6.67408 * (10 ^ -11)
```

## The Data

I would like to use three of the datasets from The Star Wars API (SWAPI): (1) the characters, (2) the planets, and (3) the starships. Fortunately, the `repurrrsive` package contains the Star Wars datasets.

### The People

```{r load_person_data, warning = FALSE}
# load and format data
people <- data.frame(t(list.cbind(sw_people))) %>%
  as.tibble() %>%
  select(name:gender) %>%
  mutate_if(is.list, ~(unlist(.))) %>%
  mutate_all(~(na_if(., "unknown"))) %>%
  mutate_at(vars(height,
                 mass), 
            ~(as.numeric(.))) %>%
  mutate(birth_year = parse_number(as.character(birth_year)),
         hair_color = if_else(gender == "n/a", "droid", hair_color),
         gender     = if_else(gender == "n/a", "droid", gender))

# look at the data
head(people)
```


### The Starships

```{r load_starships_data, warning = FALSE}
# load and format data
starships <- data.frame(t(list.cbind(sw_starships))) %>%
  as.tibble() %>%
  janitor::clean_names() %>%
  select(name:starship_class) %>%
  mutate_if(is.list, ~(unlist(.))) %>%
  mutate_all(funs(na_if(., "unknown"))) %>%
  mutate(max_atmosphering_speed = parse_number(max_atmosphering_speed),
         cost_in_credits        = as.numeric(cost_in_credits),
         length                 = as.numeric(length),
         crew                   = as.numeric(crew),
         passengers             = as.numeric(passengers),
         cargo_capacity         = as.numeric(cargo_capacity),
         hyperdrive_rating      = as.numeric(hyperdrive_rating),
         mglt                   = as.numeric(mglt),
         starship_class         = as.factor(tolower(starship_class))) %>%
  separate(consumables, c("con_n", "con_unit"), sep = " ") %>%
  mutate(con_unit    = case_when(con_unit == "day"   ~ "1",
                                 con_unit == "week"  ~ "7",
                                 con_unit == "month" ~ "30",
                                 con_unit == "years" ~ "365"),
         consumables = as.numeric(con_n) * as.numeric(con_unit)) %>%
  select(-starts_with("con_")) %>%
  select(which(sapply(., is.character)), 
         which(sapply(., is.factor)), 
         which(sapply(., is.numeric)))

# look at data
head(starships) 
```


### The Planets

```{r load_planets_data, warning = FALSE}
# load and format data
planets <- data.frame(t(list.cbind(sw_planets))) %>%
  as.tibble() %>%
  select(name:population) %>%
  mutate_if(is.list, ~(unlist(.))) %>%
  mutate_all(~(na_if(., "unknown"))) %>%
  mutate_at(vars(rotation_period,
                 orbital_period,
                 diameter,
                 surface_water,
                 population),
            ~(as.numeric(.))) %>%
  mutate(surface_water   = surface_water / 100,
         population_mils = population / 1000000,
         au              = (((orbital_period / 365) ^ 2) ^ (1/3)), # calculate distance from sun accoridng to Kepler's 3rd law
         gravity         = parse_number(as.character(gravity)),
         g               = gravity * 9.8, # calculate force of gravity
         radius          = diameter / 2, # calculate radius
         mass            = (g * (radius^2)) / grav_constant, #calculate planet mass
         area            = pi * (radius^2)) %>% # calculate planet area
  select(which(sapply(., is.character)), which(sapply(., is.numeric)))

# look at the data
head(planets)
```


## The Visualizations

### The People

```{r visualization_1}
# people plot
people %>%
  filter(gender == "male" | gender == "female" | gender == "droid") %>%
  group_by(gender) %>%
  summarise(m_age = mean(birth_year, na.rm = TRUE)) %>%
  ggplot(aes(x = factor(gender), y = m_age, fill = factor(gender))) +
    geom_col(width = 0.10) +
    geom_col(width = 0.12, alpha = .6) +
    geom_col(width = 0.14, alpha = .4) +
    geom_col(width = 0.16, alpha = .2) +
    geom_col(width = 0.18, alpha = .1) +
    scale_fill_manual(values = c("#0FFF00", "#FF0000", "#000CFF")) +
    geom_rect(aes(ymin = -10, 
                  ymax = 0, 
                  xmin = .95,
                  xmax = 1.05),
              fill = "gray30") +
    geom_rect(aes(ymin = -10, 
                  ymax = 0, 
                  xmin = 1.95,
                  xmax = 2.05),
              fill = "gray30") +
    geom_rect(aes(ymin = -10, 
                  ymax = 0, 
                  xmin = 2.95,
                  xmax = 3.05),
              fill = "gray30") +
    geom_rect(aes(ymin = -5, 
                  ymax = -20, 
                  xmin = .94,
                  xmax = 1.06),
              fill = "gray30") +
    geom_rect(aes(ymin = -5, 
                  ymax = -20, 
                  xmin = 1.94,
                  xmax = 2.06),
              fill = "gray30") +
    geom_rect(aes(ymin = -5, 
                  ymax = -20, 
                  xmin = 2.94,
                  xmax = 3.06),
              fill = "gray30") +
    theme_bw() +
    labs(title = "Average age of Star Wars characters for women, men, and droids at the Battle of Yavin",
           y = "Age (years)",
           x = "Gender",
           legend = "Gender",
           caption = "Data from https://swapi.co.")
```

#### Idea

Although not a great data/ink ratio, I believe the above figure would be memorable in displaying the ages of the characters in Starwars. I would like to add more design elements to the plot (e.g., a more detailed lightsaber hilt, a glow on top of the saber). It would also be good to add some text labels showing the actual ages of characters beside the bars. Yoda might also be an outlier, so it may be appropriate to eliminate him from the plot. 

#### Intended Audience

The intended audience for this visualization are fans of Star Wars and the general public.

#### Intended Message

Other than wanting to convey the average difference in ages among genders in the Star Wars universe, I think it is important to consider that the mean age of female characters is so young compared to that of men. The recent movies have begun to include older women (e.g., Laura Dern), but this is partially due to the aging of actors that were in the original films (i.e., Carrie Fisher). Star Wars would benefit from including an array of ages in women akin to that in men.

```{r visualization_2}
# starships plot
starships %>%
  group_by(starship_class) %>%
  filter(!is.na(length),
         n() > 1) %>%
  summarize(len_m = mean(length),
            len_sd = sd(length)) %>%
  filter(!is.na(len_sd)) %>%
    ggplot(aes(x = fct_reorder(starship_class, desc(len_m)), y = len_m)) +
      geom_errorbar(aes(ymin = len_m - len_sd, ymax = len_m + len_sd), width = .5) +
      geom_point(size = 5) +
      geom_point(size = 3, shape = 13, color = "grey80") +
      coord_flip() +
      theme_bw() +
      labs(title = "Average length of starship by ship class",
           y = "Average Length (meters)",
           x = "Ship class",
           caption = "Data from https://swapi.co.")
```

#### Idea

Points and standard error bars have always looked like Tie Fighters to me, so I decided I would try to plot the length of a starship against the starship class. As would be expected yachts are the largest ships with assult starfighters being slightly larger than basic starfighters. Although I am quite happy with the plot, I believe the fonts and text elements could be greatly improved. Representing all ships as tie fighters also may be misleading. 

#### Intended Audience

The intended audience for this visualization again are people who like starwars but to appreciate the error bars some knowledge of statistics would likely be needed. 

#### Intended Message

That starship lengths differ by class and that the length of a yacht starship is far more variable than other starship classes. 


```{r visualization_3}
# planets plot
planets %>%
  filter(area < mean(area, na.rm = TRUE) + (sd(area, na.rm = TRUE) * 3)) %>%
  ggplot(aes(x = au, y = surface_water, size = area)) +
    geom_point(alpha = .8, color = "yellow") +
    geom_smooth(method = "lm", se = FALSE, color = "yellow") +
    scale_color_gradient(high = "#0E0584", low = "#086703", na.value = "#AB8703") +
    theme(panel.background = element_rect(fill = "black"),
          panel.grid.major = element_line(size = .10),
          panel.grid.minor = element_line(size = .05)) +
    scale_y_continuous(labels = scales::percent) +
    xlim(.5,1.5) +
    labs(title = "Distance from the sun and amount of surface water",
         y = "Amount of surface water (%)",
         x = "Distance from star (Astronomical Units)",
         size = "Size",
         caption = "Data from https://swapi.co")

```

#### Idea

As interesting as I found the data in the first two plots and after tinkering around with the planets dataset, I think several plots could be devoted to this one dataset. Especially, given the ability of the visualizations to be framed around teaching astronomy (e.g., Keppler's third law, calculations of gravitational force). The present plot shows the relationship between a planet's surface water and its distance from the sun.

#### Intended Audience

Fans of Star Wars who are interested in astronomy. 

#### Intended Message

Planets farther from the sun seem to have more surface water, which seems to be consistent with our solar system. 
