---
title: "edld610_catd_finalproject"
author: "Cameron Kay"
date: "2/3/2019"
output:
  html_document:
    highlight: kate
    theme: cosmo
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r initial_setup, include = FALSE}
# load necessary packages (if you do not have them, you will need to install them first)
library(tidyverse) # for dataframe manipulation and plotting
library(magrittr) # for updating an object that is piped into a function (using `%<>%`)
library(here) # for finding the project folder
library(repurrrsive) # contains the star wars datasets
library(rlist) # for binding all lit elements by column (using `list.cbind`)
library(ggrepel) # for adding geom_text labels that do not overlap

# add gravitational constant (i.e., 6.67408 * 10^-11)
grav_constant <- 6.67408 * (10 ^ -11)

# load bespoke functions from the functions folder
source(here("functions/my_theme.R")) # load my_theme
source(here("functions/geom_saberhilt.R")) # load the function for creating the hilts of lightsabers
source(here("functions/geom_saberblade.R")) # load the function for creating the blades of lightsabers

# set figure output dimensions
knitr::opts_chunk$set(fig.width  = 12, 
                      fig.height = 8)
```

# AE

I really like the way you use annotations (#) in your code to specify what all of your commands are doing each step of the way. It makes it super easy for someone not using or familiar with your dataset to keep up and understand your data manipulations and visualizations. This is a practice I'd like to incorporate more into my own coding. You've even specified what all your loaded packages are being used to do. Super mindful!

## The Data

### The People

```{r load_person_data, warning = FALSE}
# load and format data
people <- data.frame(t(list.cbind(sw_people))) %>%
  # convert the star wars people dataframe to a tibble
  as_tibble() %>%
  # select only the columns between name and gender (inclusive)
  select(name:gender) %>%
  # mutates columns that are lists to non-lists
  mutate_if(is.list, ~(unlist(.))) %>%
  # mutate all columns and replaces the text "unknown" with "NA"
  mutate_all(~(na_if(., "unknown"))) %>%
  # converts height and mass to numeric variables
  mutate_at(vars(height, mass), ~(as.numeric(.))) %>%
  # extracts birth year and converts it to numeric; replace "n/a" values with 
  # droid for hair_color and gender
  mutate(birth_year = parse_number(as.character(birth_year)),
         hair_color = if_else(gender == "n/a", "droid", hair_color),
         gender     = if_else(gender == "n/a", "droid", gender))

# look at first ten rows of tidied dataet
head(people, n = 10L)
```

# AE

Even your spacing in the `if_else` statement above is excellent. I like that everything on the right side of the equals sign is aligned at the same position on its respective line.

### The Starships

```{r load_starships_data, warning = FALSE}
# load and format data
starships <- data.frame(t(list.cbind(sw_starships))) %>%
  # convert the star ways starship dataset to a tibble
  as_tibble() %>%
  # clean column names
  janitor::clean_names() %>%
  # select only those columns between name and starship (inclusive)
  select(name:starship_class) %>%
  # mutate columns that are lists to non-lists
  mutate_if(is.list, ~(unlist(.))) %>%
  # mutate all columns, replacing "unknown" with "NA"  
  mutate_all(~(na_if(., "unknown"))) %>%
  # convert numeric columns to numerics
  mutate_at(vars(cost_in_credits,
                 length,
                 crew,
                 passengers,
                 cargo_capacity,
                 hyperdrive_rating,
                 mglt), ~(as.numeric(.))) %>%
  # extract the max_atmosphering_speed and convert starship_class to a lowercase
  # factor
  mutate(max_atmosphering_speed = parse_number(max_atmosphering_speed),
         starship_class         = as.factor(tolower(starship_class))) %>%
  # seperate consumables into a value and unit (e.g., days, weeks, months, 
  # years)
  separate(consumables, c("con_n", "con_unit"), sep = " ") %>%
  # convert units to numeric values and multiply with the consumable value to 
  # provide the amount of consumable in the total number of days
  mutate(con_unit    = case_when(con_unit == "day"   ~ "1",
                                 con_unit == "week"  ~ "7",
                                 con_unit == "month" ~ "30",
                                 con_unit == "years" ~ "365"),
         consumables = as.numeric(con_n) * as.numeric(con_unit)) %>%
  # retain only the new consumables column (i.e., drop con_n and con_unit)
  select(-starts_with("con_"))

# look at first ten rows of tidied dataet
head(starships, n = 10L) 
```

### The Planets

```{r load_planets_data, warning = FALSE}
# load and format data
planets <- data.frame(t(list.cbind(sw_planets))) %>%
  # convert star wars planets dataset to a tibble
  as.tibble() %>%
  # retain only columns between name and population (inclusive)
  select(name:population) %>%
  # mutate columns that are lists to non-lists
  mutate_if(is.list, ~(unlist(.))) %>%
  # mutate all columns, replacing "unknown" with "NA"  
  mutate_all(~(na_if(., "unknown"))) %>%
  # convert numeric columns to type numeric
  mutate_at(vars(rotation_period,
                 orbital_period,
                 diameter,
                 surface_water,
                 population),
            ~(as.numeric(.))) %>%
  # calculate values of interest
  mutate(
         # make surface water a proportion
         surface_water   = surface_water / 100,
         # calculate population in millions
         population_mils = population / 1000000,
         # calculate a planet's distance from the sun (Kepler's 3rd law)
         au              = (((orbital_period / 365) ^ 2) ^ (1/3)),
         # convert gravity to a number
         gravity         = parse_number(as.character(gravity)),
         # calculate the force of gravity
         g               = gravity * 9.8, 
         # calculate radius
         radius          = diameter / 2, 
         # calculate planet mass (ish)
         mass            = (g * (radius^2)) / grav_constant,
         # calculate planet area
         area            = pi * (radius^2))

# look at first ten rows of tidied dataet
head(planets, n = 10L) 
```

# AE

For your final portfolio submission, I would add some information to contextualize the visualizations you provide here. While your code does a good job explaining each step you're taking to massage your data into its desired format, I don't currently have a good understanding of what you are seeking to show through the data. 

## The Visualizations

### The People

```{r visualization_1}
# people plot
(people2 <- people %>%
  # filter for only females and males
  filter(gender == "female" | gender == "male") %>%
  # group by gender
  group_by(gender) %>%
  # calculate mean age and rounded mean age for geom_text
  summarise(m_age   = mean(birth_year, na.rm = TRUE),
            m_age_r = round(m_age, 2),
            m_age_r = paste(m_age_r, "YEARS")) %>%
  # convert genders to title case and gender to a factor
  mutate(gender = str_to_title(gender),
         gender = as.factor(gender)) %>%
  # plot gender vs age and fill by gender
  ggplot(aes(x = gender, y = m_age, fill = gender)) +
    # add the lightsaber blades to the plot
    geom_saberblade() +
    # add the lightsaber hilts to the plot
    geom_saberhilt(column = 1) +
    geom_saberhilt(column = 2) +
    # set colour and fill values (i.e., red and green)
    scale_fill_manual(values   = c("#0FFF00", "#00f6ff")) +
    scale_colour_manual(values = c("#0FFF00", "#00f6ff")) +
    # add text with the mean ages to the plot
    geom_text(aes(label = m_age_r, color = gender), 
              size     = 6, 
              position = position_nudge(x = -.09, y = -8)) +
    # add labels
    labs(title     = "Average age and gender",
           y       = "Age (years)",
           x       = "Gender",
           legend  = "Gender",
           caption = "Data from https://swapi.co.") +
    # customize using my theme
    my_theme() +
    # flip the coordinates
    coord_flip())

```

# AE

The code chunk above presented an error that said "column 'gender' can't be used as a grouping variable because it's a list." To remedy this, I assigned your visualization to a new object I called people2. That seemed to do the trick, as it returned your visualization with no errors. Whew, an easy fix!

As for the visualization itself, I'm a big fan of the lightsaber aesthetic! However, I don't know if this is the most interesting or illustrative visualization, given your dataset. First, it's unclear from the visualization alone where you are drawing your population. I know that this is a comparison of male and female characters in Star Wars, but this information is not present in the visualization itself. I would indicate that in the title: "Average age of all Star Wars characters by gender"? At the very least, I would change the title from the one you currently have. You aren't presenting your audience with average age *and* gender. You're giving them age data grouped *by* gender.

It's interesting that there is what I assume to be a significant difference in average age between male and female Star Wars characters. This seems to be driven mainly by a major outlier: Yoda, the 896 year-old Jedi master! Such information could be informative and entertaining to show. You might actually want to complement the mean age visualization with another visualization that depicts the distribution of raw ages of each character by gender. The "Yoda effect" makes for a hilarious and unique takeaway from your data.

Finally, this might be a real knit-pick, but I'm not the biggest fan of the black and hot pink color palette you've used around your plot. The bolded font of the labels is also a bit hard to look at. I would experiment with the palette and labels a bit. 

### The Starships
```{r visualization_2}
# generate random x, y, and transparency data to use for plotting stars
stars_data <- tibble(x     = sample(seq(from =  0, to =  4, by = .0001), 500),
                     y     = sample(seq(from =  0, to = 80, by = .0001), 500),
                     trans = sample(seq(from = .1, to = .9, by = .0001), 500))

# starships plot
starships %>%
  # group by starship_class
  group_by(starship_class) %>%
  # drop rows that have an NA for length 
  filter(!is.na(length)) %>%
  # calculate mean length and standard deviation, as well as number of ships in 
  # that class
  summarize(len_m  = mean(length),
            len_sd = sd(length),
            len_n  = n()) %>%
  # retain classes that have more than one ship
  filter(len_n > 1) %>%
  # convert starship_class to title case and wrap multi-word levels
  mutate(starship_class = str_to_title(starship_class),
         starship_class = gsub(" ", " \n", starship_class)) %>%
  # calculate standard error
  mutate(len_se = len_sd / sqrt(len_n)) %>%
  # plot starship class v starship length
  ggplot(aes(x = fct_reorder(starship_class, desc(len_m)), y = len_m)) +
    # add error bars
    geom_errorbar(aes(ymin = len_m - (len_se * qnorm(.975)), 
                      ymax = len_m + (len_se * qnorm(.975))), 
                  width = .5, colour = "grey70", lwd = 1.25) +
    geom_errorbar(aes(ymin = len_m - (len_se * qnorm(.975)), 
                      ymax = len_m + (len_se * qnorm(.975))), 
                  width = .5, colour = "grey30", lwd = 1) +
    # add stars 
    geom_point(data = stars_data, aes(x = x, y = y, alpha = trans), 
               colour = "white",
               size = .001) +
    # plot ship body points
    geom_point(size = 6.25, colour = "grey70") +
    geom_point(size = 6.00, colour = "grey30") +
    geom_point(size = 3.00, colour = "grey10") +
    geom_point(size = 4.00, colour = "grey60", shape = 13) +
    # flip the coordinates
    coord_flip() +
    # add labels
    labs(title   = "Average length of starship by ship class",
         y       = "Average Length in Meters",
         x       = "Class",
         caption = "Data from https://swapi.co.") + 
    # set y-axis limits and do not show value beyond limits of y-axis
    scale_y_continuous(limits = c(0,80), 
                       expand = expand_scale(mult = c(0, 0))) + 
    # customize using my theme
    my_theme()
```

# AE

This visualization is *super* creative! What a memorable plot. The repurposing of starships as error bars is absolutely brilliant. Fantastic work, dude. Using a randomized distribution of white points to put stars in the background is also *way* cool. Very impressed by this.

Again, just make sure your title indicates with clarity what your audience is seeing. Be sure to let people know these are from Star Wars. Also, I wonder if it would be neat to specify what Episodes each of three starships appeared in... 

Finally, the bolded hot pink still bothers me. The inside of the plot looks so great, but I feel the outside comprises it.

### The Planets
```{r visualization_3, fig.height = 12, warning = FALSE}
# planets plot
planets %>%
  # select only variables of interest
  select(name, au, radius) %>%
  # drop planets that do not have (1) a name, (2) a distance from the sun (i.e.
  # an AU), (3) an AU less than 2, and (4) a radius greater than 0.
  filter(!is.na(name),
         !is.na(au),
         # au < 2,
         radius != 0.00) %>%
  # create a new variable indicating whether a planet was in the original 
  # trilogy
  mutate(orig_tril = case_when(name == "Alderaan" |
                               name == "Tatooine" |
                               name == "Yavin"    |
                               name == "Yavin 4"  ~ "new_hope",
                               name == "Bespin"   |
                               name == "Dagobah"  |                                 
                               name == "Hoth"     ~ "empire_strikes",                                
                               name == "Endor"    ~ "return_jedi",
                               TRUE ~ NA_character_)) %>%
  # sort the data by their distance from the sun
  arrange(au) %>%
  # add a column that arbitrarily assigns the planet a degree
  mutate(degree = seq(from = 1, to = 360, length.out = n())) %>%
  # plot the planets distance from the sun against their arbitrary degree
  ggplot(aes(x = degree, y = au)) +
    # add labels indicating y-axis units and add star and earth
    geom_text(aes(x =   0, y = 0, label = "STAR"), 
              color = "deeppink3", 
              size  = 7) +
    geom_text(aes(x = 270, y = 2, label = "2 AU"), 
              color = "deeppink3", 
              size  = 7) +
    geom_text(aes(x = 270, y = 4, label = "4 AU"), 
              color = "deeppink3", 
              size  = 7) +
    geom_text(aes(x = 270, y = 6, label = "6 AU"), 
              color = "deeppink3", 
              size  = 7) +
    # add solar system planets
    geom_text(aes(x = 0, y = 1.00, label =   "EARTH"), 
      color = "deeppink3", 
      size  = 5) +
    geom_text(aes(x = 0, y = 1.52, label =    "MARS"), 
      color = "deeppink3", 
      size  = 5) +
    geom_text(aes(x = 0, y = 5.20, label = "JUPITER"), 
      color = "deeppink3", 
      size  = 5) +
    # add planet points; map size of the point to planet radius
    geom_point(aes(size = radius), alpha = .6, color = "#00ddff") +
    # make the coordinate polar
    coord_polar(theta = "x") +
    # set the limit of size values
    scale_size_continuous(range = c(.5, 10)) +
    # drop x-axis grid marks and breaks
    scale_x_continuous(breaks = NULL) +
    # set 2-axis limit
    scale_y_continuous(limits = c(0,7)) +
    # add labels
    labs(title    = "Distance of Star War's Planets from their Host Stars",
         subtitle = "in Astronomical Units (AU)", 
         caption  = "Data from https://swapi.co.") + 
    # customize using my theme
    my_theme() +
    # drop axis text and axis titles and center title, subtitle, and caption
    theme(plot.title    = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          plot.caption  = element_text(hjust = 0.5),
          axis.text     = element_blank(),
          axis.title    = element_blank(),
          plot.margin   = unit(c(1, 2, 1, 2), "cm")) +
    # add text identifying only those planetss from the original trilogy
    geom_label_repel(data = . %>% filter(!is.na(orig_tril)), aes(label = name),
                     color              = "#00ddff",
                     fill               =        NA,
                     size               =         8,
                     min.segment.length =         0,
                     nudge_y            =       2.1)
``` 

# AE

Again, a very cool visualization. However, this one needs a lot of work. I feel like a visualization like this almost *has* to be interactive. I instinctively started clicking and dragging on the plot to see if it would move. It's just that there is so much information being communicated in a relatively small space. There should be a way to alter your field of view when looking at this to change the visual perspective. Otherwise, there is too much visual stimuli to take in at once, which makes it difficult to glean insights from the data. More specifically, there is a lot of labeling text to indicate each of the planets. However, with each of the blue circles clumped so close together, that labeling information becomes almost meaningless. This viz has a lot of potential, but I think it requires some interactivity to work as intended. 

Additionally, I see you have AU units included in the plot. Is there a way to better translate to a general audience what an astrological unit really means? Or is this meant for a more specialist community that would understand?
