---
title: "edld610_catd_finalproject_proposal"
author: "Cameron Kay"
date: "2/3/2019"
output: html_document
---

```{r load_packages, include = FALSE}
# load necessary packages
library(tidyverse)
library(magrittr)
library(repurrrsive)
library(rlist)

# add gravitational constant
grav_constant <- 6.67408 * (10 ^ -11)
```

## The Data

I would like to use three of the datasets from The Star Wars API (SWAPI): (1) the characters, (2) the planets, and (3) the starships. Fortunately, the `repurrrsive` package contains the Star Wars datasets.

### The People

```{r load_person_data, warning = FALSE}
# load and format data
people <- data.frame(t(list.cbind(sw_people))) %>%
  as.tibble() %>%
  select(name:gender) %>%
  mutate_if(is.list, ~(unlist(.))) %>%
  mutate_all(~(na_if(., "unknown"))) %>%
  mutate_at(vars(height,
                 mass), 
            ~(as.numeric(.))) %>%
  mutate(birth_year = parse_number(as.character(birth_year)),
         hair_color = if_else(gender == "n/a", "droid", hair_color),
         gender     = if_else(gender == "n/a", "droid", gender))

# look at the data
head(people)
```

### The Planets

```{r load_planets_data, warning = FALSE}
# load and format data
planets <- data.frame(t(list.cbind(sw_planets))) %>%
  as.tibble() %>%
  select(name:population) %>%
  mutate_if(is.list, ~(unlist(.))) %>%
  mutate_all(~(na_if(., "unknown"))) %>%
  mutate_at(vars(rotation_period,
                 orbital_period,
                 diameter,
                 surface_water,
                 population),
            ~(as.numeric(.))) %>%
  mutate(population_mils = population / 1000000,
         au              = (((orbital_period / 365) ^ 2) ^ (1/3)), # calculate distance from sun accoridng to Kepler's 3rd law
         gravity         = parse_number(as.character(gravity)),
         g               = gravity * 9.8, # calculate force of gravity
         radius          = diameter / 2, # calculate radius
         mass            = (g * (radius^2)) / grav_constant) %>% #calculate planet mass
  select(which(sapply(., is.character)), which(sapply(., is.numeric)))

# look at the data
head(planets)
```

### The Starships

```{r load_starships_data, warning = FALSE}
# load and format data
starships <- data.frame(t(list.cbind(sw_starships))) %>%
  as.tibble() %>%
  janitor::clean_names() %>%
  select(name:starship_class) %>%
  mutate_if(is.list, ~(unlist(.))) %>%
  mutate_all(funs(na_if(., "unknown"))) %>%
  mutate(max_atmosphering_speed = parse_number(max_atmosphering_speed),
         cost_in_credits        = as.numeric(cost_in_credits),
         length                 = as.numeric(length),
         crew                   = as.numeric(crew),
         passengers             = as.numeric(passengers),
         cargo_capacity         = as.numeric(cargo_capacity),
         hyperdrive_rating      = as.numeric(hyperdrive_rating),
         mglt                   = as.numeric(mglt)) %>%
  separate(consumables, c("con_n", "con_unit"), sep = " ") %>%
  mutate(con_unit    = case_when(con_unit == "day"   ~ "1",
                                 con_unit == "week"  ~ "7",
                                 con_unit == "month" ~ "30",
                                 con_unit == "years" ~ "365"),
         consumables = as.numeric(con_n) * as.numeric(con_unit)) %>%
  select(-starts_with("con_")) %>%
  select(which(sapply(., is.character)), which(sapply(., is.numeric)))

# look at data
head(starships) 
```

## The Visualizations

### The People

```{r visualization_1}
# load and format data
people %>%
  filter(gender == "male" | gender == "female" | gender == "droid") %>%
  group_by(gender) %>%
  summarise(m_age = mean(birth_year, na.rm = TRUE)) %>%
  ggplot(aes(x = factor(gender), y = m_age, fill = factor(gender))) +
    geom_col(width = 0.10) +
    geom_col(width = 0.12, alpha = .6) +
    geom_col(width = 0.14, alpha = .4) +
    geom_col(width = 0.16, alpha = .2) +
    geom_col(width = 0.18, alpha = .1) +
    scale_fill_manual(values = c("#0FFF00", "#FF0000", "#000CFF")) +
    geom_rect(aes(ymin = -10, 
                  ymax = 0, 
                  xmin = .95,
                  xmax = 1.05),
              fill = "gray30") +
    geom_rect(aes(ymin = -10, 
                  ymax = 0, 
                  xmin = 1.95,
                  xmax = 2.05),
              fill = "gray30") +
    geom_rect(aes(ymin = -10, 
                  ymax = 0, 
                  xmin = 2.95,
                  xmax = 3.05),
              fill = "gray30") +
    geom_rect(aes(ymin = -5, 
                  ymax = -20, 
                  xmin = .94,
                  xmax = 1.06),
              fill = "gray30") +
    geom_rect(aes(ymin = -5, 
                  ymax = -20, 
                  xmin = 1.94,
                  xmax = 2.06),
              fill = "gray30") +
    geom_rect(aes(ymin = -5, 
                  ymax = -20, 
                  xmin = 2.94,
                  xmax = 3.06),
              fill = "gray30") +
    theme_bw() +
    labs(title = "Average age of Star Wars characters for women, men, and droids at the Battle of Yavin",
           y = "Age (years)",
           x = "Gender",
           legend = "Gender")
```

**Idea**

Although not a great data/ink ratio, I believe the above figure would be memorable in displaying the ages of the characters in Starwars. I would like to add more design elements to the plot (e.g., a more detailed lightsaber hilt, a glow on top of the saber). It would also be good to add some text labels showing the actual ages of characters beside the bars. Yoda might also be an outlier, so it may be appropriate to eliminate him from the plot. 

**Intended Audience** 

The intended audience for this visualization are fans of Star Wars and the general public.

**Intended Message**

Other than wanting to convey the average difference in ages among genders in the Star Wars universe, I think it is important to consider that the mean age of female characters is so young compared to that of men. The recent movies have begun to include older women (e.g., Laura Dern), but this is partially due to the aging of actors that were in the original films (i.e., Carrie Fisher). Star Wars would benefit from including an array of ages in women akin to that in men.

```{r visualization_2}
# load and format data
planets %>%

  ggplot(aes(x = au, y = population, size = diameter, color = surface_water)) +
    geom_point(alpha = .7) +
    scale_color_gradient(high = "#0E0584", low = "#086703", na.value = "#AB8703") +
    theme(panel.background = element_rect(fill = "black"),
          panel.grid.major = element_line(size = .10),
          panel.grid.minor = element_line(size = .05)) +
    xlim(0, 7.5) +
    ylim(0, 30) +
    labs(y = "Population (Billions)",
         x = "Distance from the sun (AUs)")
```










