---
title: "Visualizing Star Wars"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    css: resources/css/custom.css
    logo: resources/images/jedi_logo.png
    favicon: resources/images/jedi_fav.png
---

```{r setup, include = FALSE}
# load necessary packages (if you do not have them, you will need to install them first)
library(flexdashboard) # for creating the flex dashboard. 
library(DT) # for creating table html widgets
library(tidyverse) # for dataframe manipulation and plotting
library(magrittr) # for updating an object that is piped into a function (using `%<>%`)
library(here) # for finding the project folder
library(repurrrsive) # contains the star wars datasets
library(rlist) # for binding all lit elements by column (using `list.cbind`)
library(ggrepel) # for adding geom_text labels that do not overlap
library(gganimate) # for animating the plot

# add gravitational constant (i.e., 6.67408 * 10^-11)
grav_constant <- 6.67408 * (10 ^ -11)

# load bespoke functions from the functions folder
source(here("resources/functions/my_theme.R")) # load my_theme
source(here("resources/functions/geom_saberhilt.R")) # load the function for creating the hilts of lightsabers
source(here("resources/functions/geom_saberblade.R")) # load the function for creating the blades of lightsabers

# set figure output dimensions
knitr::opts_chunk$set(fig.width  = 12, 
                      fig.height = 8)
```

# The Data {.storyboard data-icon="fa-database"}

### Introducing the Data

Welcome! This portfolio shows the construction of three different visualizations 
of the Star Wars universe. The [first visualization][The People] shows the relationship between average character age and gender. The [second visualization][The Starships] reveals how starship classes differ in terms of length. The [third visualization][The Planets] shows the distance of Star War's planet from their host stars and animates their orbital periods. Planets from our solar system are included for easy comparison.

All the data is from the `repurrrsive` package. The `repurrrsive` package includes 
various datasets to assist in the teaching of R. The Star Wars dataset is 
originally from [SWAPI](https://swapi.co) (i.e., the Star Wars API).

### The People Data

```{r load_people_data}
# load and format data
people <- data.frame(t(list.cbind(sw_people))) %>%
  # convert the star wars people dataframe to a tibble
  as_tibble() %>%
  # select only the columns between name and gender (inclusive)
  select(name:gender) %>%
  # mutates columns that are lists to non-lists
  mutate_if(is.list, ~(unlist(.))) %>%
  # mutate all columns and replaces the text "unknown" with "NA"
  mutate_all(~(na_if(., "unknown"))) %>%
  # converts height and mass to numeric variables
  mutate_at(vars(height, mass), ~(as.numeric(.))) %>%
  # extracts birth year and converts it to numeric; replace "n/a" values with 
  # droid for hair_color and gender
  mutate(birth_year = parse_number(as.character(birth_year)),
         hair_color = if_else(gender == "n/a", "droid", hair_color),
         gender     = if_else(gender == "n/a", "droid", gender))

# embed the data
datatable(people)
```

*** 

In order to get the people data into the correct format for our visualization, we first made a `tibble` of the data from `repurrrsive`'s `sw_people` dataset. We selected only those columns of interest and unlisted the columns (most of [SWAPI](https://swapi.co)'s data is stored in lists of lists). We then converted the variables to the correct variable types and replaced "n/a" with more meaningful values. 

### The Starships Data

```{r load_starships_data, warning = FALSE}
# load and format data
starships <- data.frame(t(list.cbind(sw_starships))) %>%
  # convert the star ways starship dataset to a tibble
  as_tibble() %>%
  # clean column names
  janitor::clean_names() %>%
  # select only those columns between name and starship (inclusive)
  select(name:starship_class) %>%
  # mutate columns that are lists to non-lists
  mutate_if(is.list, ~(unlist(.))) %>%
  # mutate all columns, replacing "unknown" with "NA"  
  mutate_all(~(na_if(., "unknown"))) %>%
  # convert numeric columns to numerics
  mutate_at(vars(cost_in_credits,
                 length,
                 crew,
                 passengers,
                 cargo_capacity,
                 hyperdrive_rating,
                 mglt), ~(as.numeric(.))) %>%
  # extract the max_atmosphering_speed and convert starship_class to a lowercase
  # factor
  mutate(max_atmosphering_speed = parse_number(max_atmosphering_speed),
         starship_class         = as.factor(tolower(starship_class))) %>%
  # seperate consumables into a value and unit (e.g., days, weeks, months, 
  # years)
  separate(consumables, c("con_n", "con_unit"), sep = " ") %>%
  # convert units to numeric values and multiply with the consumable value to 
  # provide the amount of consumable in the total number of days
  mutate(con_unit    = case_when(con_unit == "day"   ~ "1",
                                 con_unit == "week"  ~ "7",
                                 con_unit == "month" ~ "30",
                                 con_unit == "years" ~ "365"),
         consumables = as.numeric(con_n) * as.numeric(con_unit)) %>%
  # retain only the new consumables column (i.e., drop con_n and con_unit)
  select(-starts_with("con_"))

# embed the data
datatable(starships)
```

*** 

Something something something

### The Planets Data

```{r load_planets_data, warning = FALSE}
# load and format data
planets <- data.frame(t(list.cbind(sw_planets))) %>%
  # convert star wars planets dataset to a tibble
  as.tibble() %>%
  # retain only columns between name and population (inclusive)
  select(name:population) %>%
  # mutate columns that are lists to non-lists
  mutate_if(is.list, ~(unlist(.))) %>%
  # mutate all columns, replacing "unknown" with "NA"  
  mutate_all(~(na_if(., "unknown"))) %>%
  # convert numeric columns to type numeric
  mutate_at(vars(rotation_period,
                 orbital_period,
                 diameter,
                 surface_water,
                 population),
            ~(as.numeric(.))) %>%
  # calculate values of interest
  mutate(
         # make surface water a proportion
         surface_water   = surface_water / 100,
         # calculate population in millions
         population_mils = population / 1000000,
         # calculate a planet's distance from the sun (Kepler's 3rd law)
         au              = (((orbital_period / 365) ^ 2) ^ (1/3)),
         # convert gravity to a number
         gravity         = parse_number(as.character(gravity)),
         # calculate the force of gravity
         g               = gravity * 9.8, 
         # calculate radius
         radius          = diameter / 2, 
         # calculate planet mass (ish)
         mass            = (g * (radius^2)) / grav_constant,
         # calculate planet area
         area            = pi * (radius^2),
         # create column indicating whether the planet is in the solar system
         solar_system    = rep(FALSE, n()),
         # create a new variable indicating whether a planet was in the original 
         # trilogy
         orig_tril = case_when(name == "Alderaan" |
                               name == "Tatooine" |
                               name == "Yavin"    |
                               name == "Yavin 4"  ~ "new_hope",
                               name == "Bespin"   |
                               name == "Dagobah"  |                     
                               name == "Hoth"     ~ "empire_strikes",    
                               name == "Endor"    ~ "return_jedi",
                               TRUE ~ NA_character_))

# create dataset for planets in the solar system
sol_sys <- tibble(
  name           = c("Mercury", "Venus", "Earth", "Mars", "Jupiter"),
  orbital_period = c(       88,     225,     365,    687,      4330),
  au             = c(      .39,     .72,       1,   1.52,      5.20),
  solar_system   = c(     TRUE,    TRUE,    TRUE,   TRUE,      TRUE),
  radius         = c(   2439.5,    6052,    6378,   3396,     71492)
)

#combine the star war's planets and solar system data into one dataset
planets <- bind_rows(planets, sol_sys)

# embed the data
datatable(planets) 
```

*** 

Something something something


# The People {.storyboard data-icon="fa-user-astronaut"}

### Version One

```{r people_viz_v1}
# people plot
people %>%
  filter(gender == "male" | gender == "female" | gender == "droid") %>%
  group_by(gender) %>%
  summarise(m_age = mean(birth_year, na.rm = TRUE)) %>%
  ggplot(aes(x = factor(gender), y = m_age, fill = factor(gender))) +
    geom_col(width = 0.10) +
    geom_col(width = 0.12, alpha = .6) +
    geom_col(width = 0.14, alpha = .4) +
    geom_col(width = 0.16, alpha = .2) +
    geom_col(width = 0.18, alpha = .1) +
    scale_fill_manual(values = c("#0FFF00", "#FF0000", "#000CFF")) +
    geom_rect(aes(ymin = -10, 
                  ymax = 0, 
                  xmin = .95,
                  xmax = 1.05),
              fill = "gray30") +
    geom_rect(aes(ymin = -10, 
                  ymax = 0, 
                  xmin = 1.95,
                  xmax = 2.05),
              fill = "gray30") +
    geom_rect(aes(ymin = -10, 
                  ymax = 0, 
                  xmin = 2.95,
                  xmax = 3.05),
              fill = "gray30") +
    geom_rect(aes(ymin = -5, 
                  ymax = -20, 
                  xmin = .94,
                  xmax = 1.06),
              fill = "gray30") +
    geom_rect(aes(ymin = -5, 
                  ymax = -20, 
                  xmin = 1.94,
                  xmax = 2.06),
              fill = "gray30") +
    geom_rect(aes(ymin = -5, 
                  ymax = -20, 
                  xmin = 2.94,
                  xmax = 3.06),
              fill = "gray30") +
    theme_bw() +
    labs(title = "Average age of Star Wars characters for women, men, and droids at the Battle of Yavin",
           y = "Age (years)",
           x = "Gender",
           legend = "Gender",
           caption = "Data from https://swapi.co.")
```

*** 

Something somethign something. 

### Version Two

```{r people_viz_v2}
people %>%
  filter(gender == "male" | gender == "female") %>%
  group_by(gender) %>%
  summarise(m_age = mean(birth_year, na.rm = TRUE)) %>%
  mutate(gender = str_to_title(gender),
         gender = as.factor(gender)) %>%
  ggplot(aes(x = gender, y = m_age, fill = gender)) +
    geom_saberblade() +
    scale_fill_manual(values = c("#0FFF00", "#FF0000")) +
    scale_color_manual(values = c("#0FFF00", "#FF0000")) +
    geom_saberhilt(column = 1) +
    geom_saberhilt(column = 2) +
    theme_bw() +
    labs(title = "Average age and gender",
           y = "Age (years)",
           x = "Gender",
           legend = "Gender",
           caption = "Data from https://swapi.co.") +
    theme(legend.position = "none", panel.grid.major.x =                   element_blank())
```

*** 

Something something something. 

### Version Three

```{r people_viz_v3}
# people plot
people %>%
  filter(gender == "male" | gender == "female") %>%
  group_by(gender) %>%
  summarise(m_age   = mean(birth_year, na.rm = TRUE),
            m_age_r = round(m_age, 2),
            m_age_r = paste(m_age_r, "YEARS")) %>%
  mutate(gender = str_to_title(gender),
         gender = as.factor(gender)) %>%
  ggplot(aes(x = gender, y = m_age, fill = gender)) +
    geom_saberblade() +
    scale_fill_manual(values = c("#0FFF00", "#FF0000")) +
    scale_colour_manual(values = c("#0FFF00", "#FF0000")) +
    scale_size_continuous(aes(label = )) +
    geom_saberhilt(column = 1) +
    geom_saberhilt(column = 2) +
    geom_text(aes(label = m_age_r, color = gender), 
              size = 6, 
              position = position_nudge(x = -.09, y = -8)) +
    labs(title = "Average age and gender",
           y = "Age (years)",
           x = "Gender",
           legend = "Gender",
           caption = "Data from https://swapi.co.") +
    my_theme() +
    coord_flip()
```

***

Something something osmethin g. 


### Version Four

```{r people_viz_v4}
# people plot
people %>%
  # filter for only females and males and drop the outlier (yoda)
  filter(gender == "female" | gender == "male",
         name   != "Yoda") %>%
  # group by gender
  group_by(gender) %>%
  # calculate mean age and rounded mean age for geom_text
  summarise(m_age   = mean(birth_year, na.rm = TRUE),
            m_age_r = round(m_age, 2),
            m_age_r = paste(m_age_r, "YEARS")) %>%
  # convert genders to title case and gender to a factor
  mutate(gender = str_to_title(gender),
         gender = as.factor(gender)) %>%
  # plot gender vs age and fill by gender
  ggplot(aes(x = gender, y = m_age, fill = gender)) +
    # add the lightsaber blades to the plot
    geom_saberblade() +
    # add the lightsaber hilts to the plot
    geom_saberhilt(column = 1) +
    geom_saberhilt(column = 2) +
    # set colour and fill values (i.e., red and green)
    scale_fill_manual(values   = c("#0FFF00", "#00f6ff")) +
    scale_colour_manual(values = c("#0FFF00", "#00f6ff")) +
    # add text with the mean ages to the plot
    geom_text(aes(label = m_age_r, color = gender), 
              size     = 6, 
              position = position_nudge(x = -.09, y = -8)) +
    # customize y-axis
    scale_y_continuous(breaks = c(0, 20, 40, 60)) +
    # add labels
    labs(title    = "Average Age of Characters by Gender",
         subtitle = "Excluding Yoda (for obvious reasons)",
         y        = "Age (years)",
         x        = "Gender",
         legend   = "Gender",
         caption  = "Data from https://swapi.co.") +
    # customize using my theme
    my_theme() +
    # flip the coordinates
    coord_flip()
```

*** 

Something something something. 

# The Starships {.storyboard data-icon="fa-space-shuttle"}

### Version One

```{r starships_viz_v1}
# starships plot
starships %>%
  group_by(starship_class) %>%
  filter(!is.na(length),
         n() > 1) %>%
  summarize(len_m  = mean(length),
            len_sd = sd(length),
            len_n  = n()) %>%
  filter(!is.na(len_sd)) %>%
  mutate(len_se = len_sd / sqrt(len_n)) %>%
    ggplot(aes(x = fct_reorder(starship_class, desc(len_m)), y = len_m)) +
      geom_errorbar(aes(ymin = len_m - (len_se * 1.96), ymax = len_m + (len_se * 1.96)), 
                    width = .5) +
      geom_point(size = 5) +
      geom_point(size = 3, shape = 13, color = "grey80") +
      coord_flip() +
      theme_bw() +
      labs(title = "Average length of starship by ship class",
           y = "Average Length (meters)",
           x = "Ship class",
           caption = "Data from https://swapi.co.")
```

*** 

Something somethign something. 

### Version Three

```{r starships_viz_v3}
# generate random x, y, and transparency data to use for plotting stars
stars_data <- tibble(x     = sample(seq(from =  0, to =  4, by = .0001), 500),
                     y     = sample(seq(from =  0, to = 80, by = .0001), 500),
                     trans = sample(seq(from = .1, to = .9, by = .0001), 500))

# starships plot
starships %>%
  # group by starship_class
  group_by(starship_class) %>%
  # drop rows that have an NA for length 
  filter(!is.na(length)) %>%
  # calculate mean length and standard deviation, as well as number of ships in 
  # that class
  summarize(len_m  = mean(length),
            len_sd = sd(length),
            len_n  = n()) %>%
  # retain classes that have more than one ship
  filter(len_n > 1) %>%
  # convert starship_class to title case and wrap multi-word levels
  mutate(starship_class = str_to_title(starship_class),
         starship_class = gsub(" ", " \n", starship_class)) %>%
  # calculate standard error
  mutate(len_se = len_sd / sqrt(len_n)) %>%
  # plot starship class v starship length
  ggplot(aes(x = fct_reorder(starship_class, desc(len_m)), y = len_m)) +
    # add error bars
    geom_errorbar(aes(ymin = len_m - (len_se * qnorm(.975)), 
                      ymax = len_m + (len_se * qnorm(.975))), 
                  width = .5, colour = "grey70", lwd = 1.25) +
    geom_errorbar(aes(ymin = len_m - (len_se * qnorm(.975)), 
                      ymax = len_m + (len_se * qnorm(.975))), 
                  width = .5, colour = "grey30", lwd = 1) +
    # add stars 
    geom_point(data = stars_data, aes(x = x, y = y, alpha = trans), 
               colour = "white",
               size = .001) +
    # plot ship body points
    geom_point(size = 6.70, colour = "grey70") +
    geom_point(size = 6.40, colour = "grey30") +
    geom_point(size = 4.40, colour = "grey10") +
    geom_point(size = 3.40, colour = "grey60", shape = 8) +
    geom_point(size = 4.40, colour = "grey60", shape = 1) +
    geom_point(size = 2.00, colour = "grey60") +
    geom_point(size = 1.50, colour = "grey10") +
    geom_point(size = 0.20, colour = "grey60", 
               position = position_nudge(x = -0.09, y = -0.5)) +
    geom_point(size = 0.20, colour = "grey60", 
             position = position_nudge(x = -0.09, y = 0.5)) +

    geom_point(size = 0.10, colour = "grey10", 
               position = position_nudge(x = -0.09, y = -0.5)) +
    geom_point(size = 0.10, colour = "grey10", 
             position = position_nudge(x = -0.09, y = 0.5)) +
    # flip the coordinates
    coord_flip() +
    # add labels
    labs(title    = "Average Length of Starship by Ship Class",
         subtitle = "Mean length indicated by ship body position\nConfidence interval (95%) indicated by wing span",
         y        = "Average Length in Meters",
         x        = "Class",
         caption  = "Data from https://swapi.co.") + 
    # set y-axis limits and do not show value beyond limits of y-axis
    scale_y_continuous(limits = c(0,80), 
                       expand = expand_scale(mult = c(0, 0))) + 
    # customize using my theme
    my_theme() +
    theme(plot.margin  = unit(c(1, 1, 1, 1), "cm"),
          axis.title.y = element_text(vjust = 5))
```

*** 

Something something something.

# The Planets {.storyboard data-icon="fa-globe-asia"}

### Version One

```{r planets_viz_v1, warning = FALSE}
# planets plot
planets %>%
  filter(area < mean(area, na.rm = TRUE) + (sd(area, na.rm = TRUE) * 3)) %>%
  ggplot(aes(x = au, y = surface_water, size = area)) +
    geom_point(alpha = .8, color = "yellow") +
    geom_smooth(method = "lm", se = FALSE, color = "yellow") +
    scale_color_gradient(high = "#0E0584", low = "#086703", na.value = "#AB8703") +
    theme(panel.background = element_rect(fill = "black"),
          panel.grid.major = element_line(size = .10),
          panel.grid.minor = element_line(size = .05)) +
    scale_y_continuous(labels = scales::percent) +
    xlim(.5,1.5) +
    labs(title = "Distance from the sun and amount of surface water",
         y = "Amount of surface water (%)",
         x = "Distance from star (Astronomical Units)",
         size = "Size",
         caption = "Data from https://swapi.co")

```

*** 

Something somethign something. 

### Version Three

```{r planets_viz_v3, fig.height = 12, warning = FALSE}
# create function for calculating x axis locations during the time it takes
# bespin to make one rotation. There has to be enough slices to obscure the
# geom_points jumping back at the end of Bespin's year
orbit_slice <- function(planet = NULL) {
  # grab orbital period of planet of interest
  temp <- planets$orbital_period[which(planets$name == planet)]
  # calculate 511 x-axis point for that planet throughout Bespin' year
  rep(seq(from         = 0, 
            to         = 5110, 
            length.out = (5110 * (temp / 5110))),
      length.out = 5110)
}

# calculate locations on the x-axis for each planets during the time it takes
# bespin to make one rotation. 
orbital_degrees <- tibble(
  "Alderaan"   = orbit_slice("Alderaan"),
  "Hoth"       = orbit_slice("Hoth"),
  "Dagobah"    = orbit_slice("Dagobah"),
  "Bespin"     = orbit_slice("Bespin"),
  "Endor"      = orbit_slice("Endor"),
  "Tatooine"   = orbit_slice("Tatooine"),
  "Mercury"    = orbit_slice("Mercury"),
  "Venus"      = orbit_slice("Venus"),
  "Earth"      = orbit_slice("Earth"),
  "Mars"       = orbit_slice("Mars"),
  "Jupiter"    = orbit_slice("Jupiter"),
  "days_count" = 1:5110
)

# gather days into a column and the planet names into another column
orbital_degrees %<>%
  gather(key = "name", value = "days", -days_count)

# planets plot
planets_plot <- planets %>%
  # select only variables of interest
  select(name, au, radius, orbital_period, orig_tril, solar_system) %>%
  # drop planets that do not have (1) a name, (2) a distance from the sun (i.e.
  # an AU), (3) an AU less than 2, and (4) a radius greater than 0.
  filter(!is.na(name),
         !is.na(au),
         !is.na(orbital_period),
         !is.na(orig_tril) | solar_system == TRUE,
         # au < 2,
         radius != 0.00) %>%
  left_join(orbital_degrees, by = "name") %>%
  #drop mercury and alderaan
  filter(name != "Mercury", # mercury for orbiting too quickly
         name != "Alderaan") %>% # alderaan for esssentially being earth 2.0
  # plot the planets distance from the sun against their arbitrary degree
  ggplot(aes(x = days, y = au, color = solar_system)) +
    # add labels indicating y-axis units and add star and earth
    geom_text(aes(x =   0,          y = 0, label = "STAR"), 
              color = "deeppink3", 
              size  = 5) +
    geom_text(aes(x = 5110 * 3 / 4, y = 2, label = "2 AU"), 
              color = "deeppink3", 
              size  = 5) +
    geom_text(aes(x = 5110 * 3 / 4, y = 4, label = "4 AU"), 
              color = "deeppink3", 
              size  = 5) +
    geom_text(aes(x = 5110 * 3 / 4, y = 6, label = "6 AU"), 
              color = "deeppink3", 
              size  = 5) +
    # add label for describing 1 AU
    annotate(geom  =                   "label", 
             x     =             5110 * 1 / 2,
             y     =                        7,
             label = "1 AU = ~150,000,000 km", 
             color =              "deeppink2",
             fill  =                  "black",
             size  =                        5) +
    # add planet points; map size of the point to planet radius
    geom_point(aes(size = radius), alpha = .6) +
    # add planet text; also suppress its legend
    geom_text(aes(label   = name), 
              position    = position_nudge(y = 1.1), 
              alpha       = .7,
              size        =  5,
              show.legend = FALSE) +
    # make the coordinate polar
    coord_polar(theta = "x") +
    # add transition according to days
    transition_manual(days_count) +
    # set the limit of size values
    scale_size_continuous(range = c(.5, 10)) +
    # drop x-axis grid marks and breaks
    scale_x_continuous(breaks = NULL) +
    # set 2-axis limit
    scale_y_continuous(limits = c(0, 7)) +
    # set the colours
    scale_colour_manual(name   = "Star System",
                        values = c("#02fff6", "#ff7b00"),
                        labels = c("Solar System", "Star Wars")) +
    # add labels
    labs(title    = "Distance from Host Star and Orbital Period",
         subtitle =                         "Days: {current_frame}", 
         caption  =                "Data from https://swapi.co.") +
    # customize using my theme
    my_theme() +
    # drop axis text and axis titles and center title, subtitle, and caption
    theme(plot.title      = element_text(hjust = 0.5),
          plot.subtitle   = element_text(hjust = 0.5),
          plot.caption    = element_text(hjust = 0.5),
          axis.text       = element_blank(),
          axis.title      = element_blank(),
          plot.margin     = unit(c(1, 5, 1, 7), "cm"),
          legend.position = "right") +
    # drop radius legend and change the size of the legend points
    guides(size = FALSE,
           colour = guide_legend(override.aes = list(size  = 5,
                                                     alpha = 1)))

# animate plot with a duration of 55 and a fps of  48
#animate(planets_plot, 
#        nframes = 55,
#        fps     = 1)

#2640
#48
``` 


*** 

Something something something.