# The Planets {#planets}

### The Data

```{r load_planets_data, warning = FALSE}
# load and format data
planets <- data.frame(t(list.cbind(sw_planets))) %>%
  # convert star wars planets dataset to a tibble
  as.tibble() %>%
  # retain only columns between name and population (inclusive)
  select(name:population) %>%
  # mutate columns that are lists to non-lists
  mutate_if(is.list, ~(unlist(.))) %>%
  # mutate all columns, replacing "unknown" with "NA"  
  mutate_all(~(na_if(., "unknown"))) %>%
  # convert numeric columns to type numeric
  mutate_at(vars(rotation_period,
                 orbital_period,
                 diameter,
                 surface_water,
                 population),
            ~(as.numeric(.))) %>%
  # calculate values of interest
  mutate(
         # make surface water a proportion
         surface_water   = surface_water / 100,
         # calculate population in millions
         population_mils = population / 1000000,
         # calculate a planet's distance from the sun (Kepler's 3rd law)
         au              = (((orbital_period / 365) ^ 2) ^ (1/3)),
         # convert gravity to a number
         gravity         = parse_number(as.character(gravity)),
         # calculate the force of gravity
         g               = gravity * 9.8, 
         # calculate radius
         radius          = diameter / 2, 
         # calculate planet mass (ish)
         mass            = (g * (radius^2)) / grav_constant,
         # calculate planet area
         area            = pi * (radius^2))

# look at first ten rows of tidied dataet
head(planets, n = 10L) 
```


### The Visualization
```{r visualization_3, fig.height = 12, warning = FALSE}
# planets plot
planets %>%
  # select only variables of interest
  select(name, au, radius) %>%
  # drop planets that do not have (1) a name, (2) a distance from the sun (i.e.
  # an AU), (3) an AU less than 2, and (4) a radius greater than 0.
  filter(!is.na(name),
         !is.na(au),
         # au < 2,
         radius != 0.00) %>%
  # create a new variable indicating whether a planet was in the original 
  # trilogy
  mutate(orig_tril = case_when(name == "Alderaan" |
                               name == "Tatooine" |
                               name == "Yavin"    |
                               name == "Yavin 4"  ~ "new_hope",
                               name == "Bespin"   |
                               name == "Dagobah"  |                                 
                               name == "Hoth"     ~ "empire_strikes",                                
                               name == "Endor"    ~ "return_jedi",
                               TRUE ~ NA_character_)) %>%
  # sort the data by their distance from the sun
  arrange(au) %>%
  # add a column that arbitrarily assigns the planet a degree
  mutate(degree = seq(from = 1, to = 360, length.out = n())) %>%
  # plot the planets distance from the sun against their arbitrary degree
  ggplot(aes(x = degree, y = au)) +
    # add labels indicating y-axis units and add star and earth
    geom_text(aes(x =   0, y = 0, label = "STAR"), 
              color = "deeppink3", 
              size  = 7) +
    geom_text(aes(x = 270, y = 2, label = "2 AU"), 
              color = "deeppink3", 
              size  = 7) +
    geom_text(aes(x = 270, y = 4, label = "4 AU"), 
              color = "deeppink3", 
              size  = 7) +
    geom_text(aes(x = 270, y = 6, label = "6 AU"), 
              color = "deeppink3", 
              size  = 7) +
    # add solar system planets
    geom_text(aes(x = 0, y = 1.00, label =   "EARTH"), 
      color = "deeppink3", 
      size  = 5) +
    geom_text(aes(x = 0, y = 1.52, label =    "MARS"), 
      color = "deeppink3", 
      size  = 5) +
    geom_text(aes(x = 0, y = 5.20, label = "JUPITER"), 
      color = "deeppink3", 
      size  = 5) +
    # add planet points; map size of the point to planet radius
    geom_point(aes(size = radius), alpha = .6, color = "#00ddff") +
    # make the coordinate polar
    coord_polar(theta = "x") +
    # set the limit of size values
    scale_size_continuous(range = c(.5, 10)) +
    # drop x-axis grid marks and breaks
    scale_x_continuous(breaks = NULL) +
    # set 2-axis limit
    scale_y_continuous(limits = c(0,7)) +
    # add labels
    labs(title    = "Distance of Star War's Planets from their Host Stars",
         subtitle = "in Astronomical Units (AU)", 
         caption  = "Data from https://swapi.co.") + 
    # customize using my theme
    my_theme() +
    # drop axis text and axis titles and center title, subtitle, and caption
    theme(plot.title    = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          plot.caption  = element_text(hjust = 0.5),
          axis.text     = element_blank(),
          axis.title    = element_blank(),
          plot.margin   = unit(c(1, 2, 1, 2), "cm")) +
    # add text identifying only those planetss from the original trilogy
    geom_label_repel(data = . %>% filter(!is.na(orig_tril)), aes(label = name),
                     color              = "#00ddff",
                     fill               =        NA,
                     size               =         8,
                     min.segment.length =         0,
                     nudge_y            =       2.1)
``` 

#### Idea

As interesting as I found the data in the first two plots, I think several plots could be devoted to the planets dataset. Specifically, the visualizations could be framed around the common theme of astronomy, serving as an instructive tool for teaching concepts from astronomy (e.g., Keppler's third law, calculations of gravitational force). The present plot shows the relationship between a planet's surface water and its distance from the sun. I would like to add a solar system dataset to the Starwar's planets dataset and use facetwrap to provide a side by side comparison of fictional and non-fictional planets. 

#### Intended Audience

Fans of Star Wars or anyone who has an interest in astronomy or phyics.

#### Intended Message

Planets farther from the sun seem to have more surface water, which seems to be consistent with our solar system. 
